% !TEX root = ../my-thesis.tex
%
\pdfbookmark[0]{Acknowledgement}{Acknowledgements}
\addchap*{Acknowledgement\label{sec:acknowledgement}}

First and foremost, I would like to express my gratitude to Günter for giving me the opportunity to pursue a PhD under his guidance
and learn from his invaluable experience and knowledge. I really appreciate your ongoing support and encouragement
throughout this journey, both academically and personally. 

A special thank you also goes to Sepp, who grew an incredible research group and environment over the past two decades.
This work would not have been possible without your constant engagement and dedication. 

I am equally grateful to all the incredible colleagues I had the privilege to work alongside. My deepest thanks to
everyone at the Institute for Machine Learning for many hours of fruitful discussions and collaboration. Thanks to
Vihang P., Theresa R., Andi Mayr, Johannes L., Johannes S., Lukas G., Sebastian L., Andi Mi., Andreu V., Markus Holzleitner, Sohvi L.,
Daniel K., Andreas F., Thomas A., Angela B., Andi R., Kristina P., Johannes Ko. and many others. A special mention goes to
Philipp S. for being an amazing co-author and an even better friend.

I would also like to acknowledge our partners from Janssen - Dries, Natalia, Paulo and Jörg - for their exceptional
collaboration and for making our joint research project an enriching experience.

To Birgit and Jenny, thank you for ensuring the institute ran smoothly, for your continuous support and 
for always being in a good mood. To Herbert, thanks for keeping our GPUs up and running.

To my family and friends, thank you for your unwavering and unconditional support. Thank you for always being there for
me and making life joyful.

Last but not least, I would like to thank my partner Jordan for being so understanding of all the stresses and
frustrations that came with this thesis. Your love and encouragement mean the world to me, and I am so grateful to have
you by my side.